---
title: "EC analysis"
author: "Lauren"
date: "8/10/2018"
output: html_document
---
# Analysis of ECs provided by different methods.

I'm required to cite:

Alexander Lex, Nils Gehlenborg, Hendrik Strobelt, Romain Vuillemot, Hanspeter Pfister,
UpSet: Visualization of Intersecting Sets,
IEEE Transactions on Visualization and Computer Graphics (InfoVis '14), vol. 20, no. 12, pp. 1983–1992, 2014.
doi:10.1109/TVCG.2014.2346248

```{r}
library(UpSetR)
library(dplyr)
library(ggplot2)
```

Read/merge ECs and TSVs.

```{r}
read_tcc <- function(path) {
    ec <-read.table(paste0(path, ".ec"), sep = '\t', stringsAsFactors = FALSE)
    names(ec) <- c("Index", "EC")
    tsv <- read.table(paste0(path, ".tsv"), sep='\t')
    tsv$V2 <- NULL
    names(tsv) <- c("Index", "Count")
    table <- inner_join(ec, tsv, by = "Index")
    table$Index <- NULL
    table <- table[table$Count != 0,]
    table
}
```

Change path here to choose between simulations and Zika datasets.

```{r}
base_path <- "~/reads/human/"
#base_path <- "~/reads/human/sims/"
datasets <- c("SRR3191542", "SRR3191543", "SRR3191544", "SRR3191545")
#datasets <- c(1:10)
#datasets <- c("SRR4045930")#, "SRR4045931")
methods <- c("kallisto", "Salmon_0.8.2", "Salmon", "HISAT2", "STAR")
ecs <- lapply(datasets, function(d) lapply(methods, function(m) read_tcc(paste0(base_path, d, "/", m, "/matrix"))))
#ecs <- lapply(datasets, function(d) lapply(methods, function(m) read_tcc(paste0(base_path, d, ".", m))))

methods <- c("kallisto", "Salmon_0.8.2", "Salmon_0.11.2", "HISAT2", "STAR")

ecs <- lapply(seq(length(datasets)), function(i) { names(ecs[[i]]) <- methods; ecs[[i]] })

tail(ecs[[1]][[1]])
```

## Simple upset plots.

```{r}
to_plot <- lapply(ecs, function(e) lapply(e, function(x) x$EC))
to_plot <- lapply(seq(length(datasets)), function(j) { names(to_plot[[j]]) <- methods; to_plot[[j]] })

invisible(lapply(seq(length(datasets)), function(j) {
    pdf(file=paste0("~/reads/human/images/", datasets[[j]], ".upset.pdf"), onefile=FALSE)
    upset(fromList(to_plot[[j]]), sets=methods, keep.order=TRUE, order.by="freq")
    dev.off();
}))
```

## Not-so-simple upset plots.
Weight by counts.

```{r}
get_intersect_table <- function(tables, m) {
    temp1 <- Reduce(function(x, y) inner_join(x, y, by="EC"), tables[names(tables) %in% m])
    #  if (length(temp1) == 0) {
    #    temp1 <- data.frame(EC=character(), None=integer(), stringsAsFactors=FALSE)
    #  } else {
    #    colnames(temp1) <- c("EC", names(tables)[names(tables) %in% m])
    #  }
    colnames(temp1)[-1] <- names(tables)[names(tables) %in% m]
    temp1 <- list(Reduce(function(x, y) anti_join(x, y, by="EC"), c(list(temp1), tables[!names(tables) %in% m])))
    names(temp1) <- c(
        paste0(names(tables)[names(tables) %in% m], collapse='&'))
    temp1
}
get_intersect_mean_sum <- function(intersect_tables) {
    temp <- lapply(intersect_tables, function(t) sum(rowSums(t[-1])) / ncol(t[-1]))
    temp
}
get_normalized_intersections <- function(tables) {
    total_counts <- sum(sapply(tables, function(t) sum(t$Count)))
    #tables <- lapply(tables, function(t) data.frame(EC=t$EC, Count=t$Count / sum(t$Count) * total_counts, stringsAsFactors=FALSE))
    combs <- lapply(lapply(seq(length(methods)), function(i) combn(methods, i, simplify=FALSE)) %>% unlist(recursive=FALSE), function(m) 
        #get_intersect_mean_sum(
        get_intersect_table(tables, m))
    #)
    combs %>% unlist(recursive=FALSE)
}
```

```{r}
invisible(lapply(seq(length(datasets)), function(j) {
    pdf(file=paste0("~/reads/human/images/", datasets[[j]], ".upset.norm.tiff"), onefile=FALSE)
    upset(fromExpression(get_normalized_intersections(ecs[[j]])), order.by="freq")
    dev.off();
}))
```

Alternatively, comment out get_intersect_mean_sum and set recursive=FALSE in the last unlist to get the original intersect tables with all the counts. Then, perform black magic.

```{r}
norm_intersects <- lapply(ecs, function(d) get_normalized_intersections(d))
norm_intersects <- lapply(norm_intersects, function(d) lapply(d, function(t) {
    if (nrow(t[-1]) != 0) {
        data.frame(AvgCount=rowMeans(t[-1]), Method=paste0(names(t[-1]), collapse='&'))
    } else {
        data.frame(AvgCount=integer(), Method=character())
    }
}))
norm_intersects <- lapply(norm_intersects, function(d) d[sort.list(lapply(d, nrow) %>% unlist, decreasing=TRUE)])
norm_intersects <- lapply(norm_intersects, function(t) do.call(rbind, t))
```

Scatter plot!

```{r}
invisible(lapply(seq(length(datasets)), function(j) {
    g <- ggplot(norm_intersects[[j]], aes(x=Method, y=AvgCount)) + geom_point(size=.5) + theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())
    ggsave(paste0("~/reads/human/images", datasets[[j]], ".intersect.boxplot.tif"), device="tiff", dpi=350)
}))
```

Bar graph!

```{r}
norm_intersects_summarized <- lapply(norm_intersects, function(d) d %>% group_by(Method) %>% summarise(Count=sum(AvgCount)))

invisible(lapply(seq(length(datasets)), function(j) {
    g <- ggplot(norm_intersects_summarized[[j]], aes(x=Method, y=Count)) + geom_col(width=.6) + theme_bw() + theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(), axis.title.y=element_text(size=20, face="bold"), axis.text.y=element_text(size=20), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_blank()) + scale_y_reverse()
    ggsave(paste0("~/reads/human/images/", datasets[[j]], ".intersect.bargraph.tif"), device="tiff", width=12, height=4, dpi=350)
}))
```

## EC length histogram.

Functions to do the thing.

```{r}
get_num_tx_tables <- function(tables) {
    tables <- lapply(seq(length(tables)), function(i) { tables[[i]]$NumTX <- sapply(tables[[i]]$EC, function(x) length(strsplit(x, ",")[[1]])); tables[[i]] })
    tables <- lapply(tables, function(t) t %>% group_by(NumTX) %>% summarise(NormCount=sum(Count)))
    tables <- lapply(seq(length(tables)), function(i) {
        tables[[i]]$NormCount <- tables[[i]]$NormCount / sum(tables[[i]]$NormCount)
        tables[[i]]$Method <- methods[[i]]
        tables[[i]]
    })
    bind_rows(tables)
}

plot_num_tx_hist <- function(table) {
    table$Method <- factor(table$Method, levels=methods)
    g <- ggplot(table, aes(x=NumTX, y=NormCount, fill=Method))
    g <- g + geom_bar(stat="identity", position="dodge", width=1)
    g <- g + xlim(0, 50) + xlab("EC size") + ylab("")
    g
}
```

Table to do the thing.

```{r}
num_tx_tables <- lapply(ecs, get_num_tx_tables)
```

Plotting the thing.

```{r}
invisible(lapply(seq(length(num_tx_tables)), function(i) {
    plot_num_tx_hist(num_tx_tables[[i]]) + labs(title=paste0("Distribution of EC sizes (", datasets[[i]], ")"))
    ggsave(paste0("~/reads/human/images/", datasets[[i]], ".EC.length.hist.tif"), device="tiff", dpi=350)
}))
```

## Numbers about the thing.
kallisto has an EC of size 1046 for all four ZIKA paired-end datasets. That's not a bug in my R code here. It has a count of 1 for all datasets except SRR3191545, where it has a count of 2.

```{r}
num_tx_stats <- lapply(num_tx_tables, function(t) Reduce(rbind, lapply(methods, function(m) {
    temp <- filter(t, Method == m)
    data.frame(Method=m, Mean=weighted.mean(temp$NumTX, temp$NormCount), Max=max(temp$NumTX))
})))
names(num_tx_stats) <- datasets
num_tx_stats
Reduce(rbind, lapply(methods, function(m) { temp <- (num_tx_stats %>% melt %>% filter(variable == "Mean" & Method == m))$value; data.frame(Method = m, Mean = mean(temp), SD = sd(temp)) }))
```

This doesn't work. Halfway through writing it, I decided I didn't even want the end product. But I'm keeping it here anyway.

```{r}
num_tx_stats <- lapply(methods, function(m) {
    temp <- lapply(num_tx_tables, function(t) filter(t, Method == m))
    means <- lapply(temp, function(t) weighted.mean(t$NumTX, t$NormCount))
    max_lengths <- lapply(temp, function(t) max(t$NumTX))
})
```


## ECs in intersections.
Looking at the differences between those EC that are in kallisto/Salmon but not in STAR/HISAT2.

```{r}
join_ECs <- function(table, method1, method2) {
    temp <- inner_join(table[[method1]], table[[method2]], "EC")
    colnames(temp) <- c("EC", method1, method2)
    temp
}

temp1 <- lapply(ecs, function(x) join_ECs(x, "kallisto", "Salmon"))
temp2 <- lapply(ecs, function(x) join_ECs(x, "HISAT2", "STAR"))
kallisto_Salmon <- mapply(function(x, y) list(anti_join(x, y, "EC")), temp1, temp2)
HISAT2_STAR <- mapply(function(x, y) list(anti_join(x, y, "EC")), temp2, temp1)
rm(temp1, temp2)

head(kallisto_Salmon[[1]])
tail(kallisto_Salmon[[1]])
head(HISAT2_STAR[[1]])
tail(HISAT2_STAR[[1]])
```

Write the intersections to file as a TCC matrix (somewhat) readable by kallisto pseudoquant (after I put it through the sparse matrix debug function).

This gives me the ability to get transcript abundances in the intersection, and from there, gene abundances. If transcript and gene abundance correlations are high, chances are, these ECs have a lot of overlapping transcripts. If not, I know which genes I should take a look at in IGV.

```{r}
write_TCC <- function(fileprefix, table, count_col, ref_ec) {
    table <- merge(ref_ec, table, by="EC", all=TRUE, sort=FALSE)
    table <- table[order(table$Index),]
    temp <- data.frame(c(0:(nrow(table) - 1)), table$EC)
    write.table(temp, file=paste0(fileprefix, ".ec"), quote=FALSE, sep='\t', row.names=FALSE, col.names=FALSE)
    temp <- data.frame(c(0:(nrow(table) - 1)), table[colnames(table) == count_col])
    write.table(temp, file=paste0(fileprefix, ".full.tsv"), quote=FALSE, sep='\t', row.names=FALSE, col.names=FALSE, na="0")
    
    file_conn <- file(paste0(fileprefix, ".cells"))
    writeLines(basename(fileprefix), file_conn)
    close(file_conn)
}
```

```{r}
ref_ec <- read.table("~/human/kallisto.idx.ec", header=FALSE, sep='\t', stringsAsFactors=FALSE)
colnames(ref_ec) <- c("Index", "EC")
write_TCC("~/reads/human/sims/1.kallisto.Salmon", kallisto_Salmon[[1]], "kallisto", ref_ec)
write_TCC("~/reads/human/sims/1.HISAT2.STAR", HISAT2_STAR[[1]], "HISAT2", ref_ec)
```

Mean and standard deviation of counts.

The range of the counts is large enough that I think I'll normalize the histogram by the counts of each EC. The means indicate that the counts tend to be low, more so for HISAT2/STAR than for kallisto/Salmon.

```{r}
kallisto_Salmon_stats <- sapply(kallisto_Salmon, function(x) c(paste("Mean", mean(x$kallisto), "SD", sd(x$kallisto), "Range:", min(x$kallisto), "-", max(x$kallisto)), paste("Mean", mean(x$Salmon), "SD", sd(x$Salmon), "Range:", min(x$Salmon), "-", max(x$Salmon))))
colnames(kallisto_Salmon_stats) <- datasets
rownames(kallisto_Salmon_stats) <- c("kallisto", "Salmon")

HISAT2_STAR_stats <- sapply(HISAT2_STAR, function(x) c(paste("Mean", mean(x$HISAT2), "SD", sd(x$HISAT2), "Range:", min(x$HISAT2), "-", max(x$HISAT2)), paste("Mean", mean(x$STAR), "SD", sd(x$STAR), "Range:", min(x$STAR), "-", max(x$STAR))))
colnames(HISAT2_STAR_stats) <- datasets
rownames(HISAT2_STAR_stats) <- c("HISAT2", "STAR")

kallisto_Salmon_stats
HISAT2_STAR_stats
```

Mean and standard deviation of differences between counts.

The counts tend not to differ much between the two methods. But then, I think that's true between all the methods, for the most part.

```{r}
kallisto_Salmon_diff <- lapply(kallisto_Salmon, function(x) paste("Mean:", mean(x$kallisto - x$Salmon), "SD:", sd(x$kallisto - x$Salmon)))
names(kallisto_Salmon_diff) <- datasets

HISAT2_STAR_diff <- lapply(HISAT2_STAR, function(x) paste("Mean:", mean(x$HISAT2 - x$STAR), "SD:", sd(x$HISAT2 - x$STAR)))
names(HISAT2_STAR_diff) <- datasets

kallisto_Salmon_diff
HISAT2_STAR_diff
```

Plotting the number of transcripts per EC.

```{r}
add_num_tx_col <- function(table) {
    table$num_tx <- sapply(table$EC, function(x) length(strsplit(x, ",")[[1]]))
    table
}

kallisto_Salmon <- lapply(kallisto_Salmon, add_num_tx_col)
HISAT2_STAR <- lapply(HISAT2_STAR, add_num_tx_col)
tail(kallisto_Salmon[[1]])
tail(HISAT2_STAR[[1]])

kallisto_Salmon_hist <- lapply(kallisto_Salmon, function(x) x %>% group_by(num_tx) %>% summarise(count=sum(kallisto)))
HISAT2_STAR_hist <- lapply(HISAT2_STAR, function(x) x %>% group_by(num_tx) %>% summarise(count=sum(HISAT2)))
for (i in 1:length(datasets)) {
    kallisto_Salmon_hist[[i]]$method <- "kallisto/Salmon"
    HISAT2_STAR_hist[[i]]$method <- "HISAT2/STAR"
}

intersect_tx_count_hist <- mapply(function(x, y) list(rbind(x, y)), kallisto_Salmon_hist, HISAT2_STAR_hist)
for (i in 1:length(datasets)) {
    intersect_tx_count_hist[[i]]$count[intersect_tx_count_hist[[i]]$method == "kallisto/Salmon"] <- intersect_tx_count_hist[[i]]$count[intersect_tx_count_hist[[i]]$method == "kallisto/Salmon"] / sum(intersect_tx_count_hist[[i]]$count[intersect_tx_count_hist[[i]]$method == "kallisto/Salmon"])
    
    intersect_tx_count_hist[[i]]$count[intersect_tx_count_hist[[i]]$method == "HISAT2/STAR"] <- intersect_tx_count_hist[[i]]$count[intersect_tx_count_hist[[i]]$method == "HISAT2/STAR"] / sum(intersect_tx_count_hist[[i]]$count[intersect_tx_count_hist[[i]]$method == "HISAT2/STAR"])
}

head(intersect_tx_count_hist[[1]])
tail(intersect_tx_count_hist[[1]])

plot_intersect_hist <- function(table) {
    g <- ggplot(table, aes(x=num_tx, y=count, fill=method))
    g <- g + geom_bar(stat="identity", position="dodge", width=1)
    g
}
```

```{r}
for (i in 1:length(datasets)) {
    g <- plot_intersect_hist(intersect_tx_count_hist[[i]])
    g
    ggsave(paste0("~/reads/human/", datasets[[i]], "/", datasets[[i]], ".intersect_tx_norm.tif"), device="tiff", dpi=350)
}
```

I want to say that generally, HISAT2/STAR has larger ECs. They definitely skew further to the right. And kallisto/Salmon generally has a great many ECs with one transcript. But kallisto still has a reasonable number (though smaller) number of larger ECs.
